---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Leonardo Rueda"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=70), tidy=TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/leor9/OneDrive/Leonardo/MIDP Courses Fall 2022/R Class/EDA-Fall2022")
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1

getwd()
library(tidyverse)
library(agricolae)
library(lubridate)

NTL_LTER_Data <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

NTL_LTER_Data$sampledate <- as.Date(NTL_LTER_Data$sampledate, format = "%m/%d/%y")
NTL_LTER_Data$year4 <- as.numeric(NTL_LTER_Data$year4)

#2


mytheme <- theme_classic(base_size = 11) + theme(axis.text = element_text(color = "black"), legend.position = "right")

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> **Answer:**
H0: The mean temperature recorded during July did not change with depth across all lakes (beta equal to zero); 
Ha: The mean temperature recorded during July changes with depth across all lakes (beta different to zero)


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL_LTER_Data_filtered_1 <- NTL_LTER_Data %>%
  mutate(month=month(sampledate)) %>% filter(month==7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5

temperaturebydepth <- 
  ggplot(NTL_LTER_Data_filtered_1, aes(x = depth, y = temperature_C)) +
  xlim(0, 16) + ylim(0, 35) + geom_point(colour = "grey") + xlab("Depth") + ylab("Temperature C ") + geom_smooth(method = 'lm', se = FALSE, color = "black") + ggtitle("Temperature by depth in July from 1984 to 2016 for all Lakes")

print(temperaturebydepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> **Answer:** There is a negative relationship between temperature and depth for all lakes during July from 1984 to 2016, that is temperature decreases as depth increases. The relationship does not seem to be completely linear, because for different depths in certain boundaries we observe the same temperature.  


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

temperaturebydepth.regression <- lm(data = NTL_LTER_Data_filtered_1, temperature_C ~ depth)

summary(temperaturebydepth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> **Answer:** According to the results of the linear regression, with 9,726 degrees of freedom and a high significance (less than 0.001), almost 73% of the variability in temperature is explained by changes in depth. For every 1m change in depth, temperature decreases 1.94 degrees celsius.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9


temperature.multiple.regression <- lm(data = NTL_LTER_Data_filtered_1, temperature_C ~ year4 + daynum + depth)

step(temperature.multiple.regression)

#10

temperature.multiple.regression.best <- lm(data = NTL_LTER_Data_filtered_1, temperature_C ~ year4 + daynum + depth)

summary(temperature.multiple.regression.best)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> **Answer:** According to the AIC method, the best set of explanatory variables to predict temperature is year4, daynum and depth. This model explains 74% of the variance in temperature, which represents 1 percentage point of improvement compared to the model using only depth as explanatory variable.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

temperaturebylakes.anova <- aov(data = NTL_LTER_Data_filtered_1, temperature_C ~ lakename)
summary(temperaturebylakes.anova)

temperaturebylakes.anova2 <- lm(data = NTL_LTER_Data_filtered_1, temperature_C ~ lakename)
summary(temperaturebylakes.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> **Answer:** According to the ANOVA test, we reject the null hypothesis that the mean temperature for all Lakes is the same. This result is significant at less than 0.001.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

temperaturebydepth2 <- 
  ggplot(NTL_LTER_Data_filtered_1, aes(x = depth, y = temperature_C)) +
  xlim(0, 16) + ylim(0, 35) + geom_point(aes(colour = lakename), alpha = 0.5) + xlab("Depth") + ylab("Temperature C ") + geom_smooth(method = 'lm', se = FALSE, color = "black") + ggtitle("Temperature by depth and Lake in July from 1984 to 2016")

print(temperaturebydepth2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(temperaturebylakes.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>**Answer:** According to the results from the Tukey test, the lakes that have the same mean temperature as Peter Lake are Paul Lake and Ward Lake. On the other hand, there is no single lake whose mean temperature is statistically different from all the other lakes (although there are significant differences between some of them). 
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>**Answer:** To check if the differences in mean temperatures between Peter Lake and Paul Lake are statistically significantly, we can perform a t-test. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

NTL_LTER_Data_filtered_2 <- NTL_LTER_Data_filtered_1 %>% filter(lakename == "Crampton Lake" | lakename == "Ward Lake") %>% na.omit()

t.test.temperaturebylake <- t.test(temperature_C ~ lakename, data = NTL_LTER_Data_filtered_2)

t.test.temperaturebylake

```

>**Answer:** The result of the t-test shows that the difference between the mean temperature of Crampton Lake and Ward Lake for July is not statistically significant. The result in part 16 shows the same. 
