---
title: "Assignment 3: Data Exploration"
author: "Leonardo Rueda"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

The completed exercise is due on Sept 30th.

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup, include=FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

```


```{r,  message = FALSE}
getwd()

setwd("C:/Users/leor9/OneDrive/Leonardo/MIDP Courses Fall 2022/R Class/EDA-Fall2022")

library(tidyverse)

Neonics_dataset <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE) # ECOTOX neonicotinoid dataset 

Litter_dataset <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE) # Niwot Ridge NEON dataset for litter and woody debris

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> **Answer:** Yes, for instance some studies have found that some concentrations of this insectiside can be harmful for the population of bees. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> **Answer:** We might be interested in studying litter and woody debris  in the forest ecosystem because they play a role in carbon budgets and nutrient cycling. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> **Answer:** 
 This are three pieces of salient information about the sampling methods: 
 1) Spatial Sampling Design: Litter and fine woody debris sampling is executed at terrestrial NEON sites that contain woody vegetation >2m
tall.
 2) Along with most of NEON’s plant productivity measurements, sampling for this product occurs only in tower
plots
 3) Temporal Sampling Design: Ground traps are sampled once per year.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}

dim(Neonics_dataset)


```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}

summary(Neonics_dataset$Effect)

```

> **Answer:** The most common effects of the study are Mortality, Population and Feeding behavior. These effects could be of interest to understand how the population of insects evolve when they are exposed to this insecticide.  

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}

summary(Neonics_dataset$Species.Common.Name)


```

> **Answer:** The most studied species of insects in the dataset are the Honeybee, the Parasitic Wasp, and the Buff Tailed Bumblebee. They are of interest because they either play a key role in the politization of several plants, or they control the population of other species in the ecosystem. 

8.  Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}

class(Neonics_dataset$Conc.1..Author.)

```

> **Answer:** The variable Conc.1..Author. in the dataset Neonics_dataset is a factor because some of the observations contain non numeric values such as the symbols /, NR/ and ~ 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}

ggplot(Neonics_dataset) + geom_freqpoly(aes(x = Publication.Year), bins = 50)


```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}

ggplot(Neonics_dataset) + geom_freqpoly(aes(x = Publication.Year, colour = Test.Location), bins = 50)

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> **Answer:** The most common test locations in the period analyzed are "Lab" and "Field natural" and the relative importance of each one has changed over time. For instance, between 1990 and 2000, "Field natural" was the most common, while between 2000 and 2020 "Lab" predominated most of the time (except for a couple of years before 2010).  

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information.

```{r}

ggplot(Neonics_dataset, aes(x = Endpoint)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> **Answer:** The two most common endpoints are "NOEL" and "LOEL". "NOEL" is defined as No-observable-effect-level, with the highest dose (concentration) producing effects not significantly different from responses of controls according to the author's reported statistical test (NOEAL/NOEC), and "LOEL" is defined as the lowest-observable-effect-level, with the lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls (LOEAL/LOEC). 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}

class(Litter_dataset$collectDate)

Litter_dataset$collectDate <- as.Date(Litter_dataset$collectDate, format = "%Y-%m-%d")

class(Litter_dataset$collectDate)

unique(Litter_dataset$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

unique(Litter_dataset$plotID)

summary(Litter_dataset$plotID)

```

> **Answer:** The "unique" function returns the number of plots that were sampled at least once, in this case 12. On the other hand, the function "summary" returns the total number of plots that were sampled for each category. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}

ggplot(Litter_dataset, aes(x = functionalGroup)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter_dataset) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup))

ggplot(Litter_dataset) +
  geom_violin(aes(x = dryMass, y = functionalGroup), 
              draw_quantiles = c(0.25, 0.5, 0.75))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> **Answer:** The boxplot is a more effective visualization because this type of graph shows a box in which we find most of the observations disregarding the distribution inside the box, while the violin plot shows how they are effectively distributed. Given that litter types are distributed fairly equally, the violin plot only shows a line.  

What type(s) of litter tend to have the highest biomass at these sites?

> **Answer:** The types of litter that tend to have the highest biomass are "needles" and "mixed". 
